Supplemental Material for ``Assessing United States county-level exposure for research on tropical cyclones and human health"

### 2019-10-09 writing/ehp_figures.tex
Work on paper, use 34 knot threshold

\begin{figure}[tbhp!] \centering
\includegraphics[width=0.7\linewidth]{hurrtracks}
\caption{States and storms considered in this study. All counties in the states
        shown in this map were investigated. The lines show the paths
	of the study storms, which included all tracked storms in 1988\,--\,2015
	\ac{HURDAT2}~\parencite{landsea2013} that came within 250~\si{\kilo\metre} 
	of at least one \ac{US} county. Thicker lines show the tracks 
	of storms whose names have been retired, indicating that the storm was particularly 
	severe or had notable impacts~\parencite{retirednames}.
	}
\label{fig:hurrtracks}
\end{figure}

@@

\begin{figure}[tbhp!] \centering
\includegraphics[width=0.9\linewidth]{raincomparison} 
	\caption{Comparison of storm-associated rainfall estimates between 
	a re-analysis dataset (included as the rainfall exposure metric
	in the open-source data and included in further analysis) and ground-based
	observations in nine sample counties. The rainfall estimates  
	include rainfall from two days before to one day after the storm's closest
	approach to the county. Each small graph shows data for one sample county, 
	and each point shows one tropical storm. The number of storms 
	within each county and the average number of stations reporting rainfall during 
	the county's storms are given above each plot. Horizontal and vertical lines 
	in each plot show the threshold of 75~\si{\milli\metre} used to classify a storm 
	as ``exposed'' in further analysis (Table 1). Note that 
	x- and y-axis ranges differ by county.
	} 
\label{fig:raincomparison}
\end{figure}

\begin{figure}[tbhp!]
\centering
\includegraphics[width=0.6\linewidth]{windcomparison}
\caption{Comparison of modeled maximum sustained surface wind speeds 
	(included as the primary wind exposure metric in the open-source
	data and used in further analysis) with estimates from the \ac{HURDAT2}
	wind speed radii. Each point represents a storm, with the x-axis
	giving the percent of counties classified in the same category
	of maximum sustained surface wind speeds ($<$34~\si{\knot}; 
	34\,--\,49.9~\si{\knot}; 50\,--\,63.9~\si{\knot}; 
	$\ge$64~\si{\knot}) by both sets of wind data for that storm. The 
	color of each point indicates the extent of the storm, 
	giving the number of study counties that were exposed to 
	maximum sustained surface wind speeds of at least 34~\si{\knot}
	(based on modeled wind speeds). Estimates are shown for [all?]
	study storms since 2004, the earliest year for which 
	wind speed radii estimates are available in \ac{HURDAT2} [?].
	}
\label{fig:windcomparison}
\end{figure}

\begin{figure}[tbhp!]
\centering
\includegraphics[width=0.8\linewidth]{floodcomparison}
\caption{Comparison for a sample of study counties of NOAA Storm Events 
	listings versus five-day total streamflows at county streamflow 
	gages during tropical cyclones. Each small plot shows results for 
	one of the sample counties. Each point represents a single tropical 
	storm, and the point's position along the x-axis shows the highest 
	daily total streamflow (cubic feet per second), summed across all 
	identified streamgages in the county, for a five-day window centered 
	on the day of the storm's closest approach to the county. The y-axis 
	separates storms for which a flood event was reported in NOAA's Storm Events 
	database for the county with a start date within the five-day window 
	of the storm's closest approach to the county. The color of each point 
	gives the percent of streamflow gages in the county with a daily streamflow 
	that exceeded a threshold of flooding on any day during the five-day window.
	Comparison for a sample of study counties of NOAA Storm Events 
	listings versus five-day total streamflows at county streamflow gages during 
	tropical storms. The number of streamflow gages used in analysis are given in 
	parentheses beside the county's name in the panel title. Each point represents 
	a single tropical storm. Note that the x-axis scales differ by county, 
	depending on the number of streamflow gages and typical flow rates for each 
	gage, and are on a log-10 scale.}
\label{fig:floodcomparison}
\end{figure}

@@

\begin{figure}%[tbhp] 
\centering
\includegraphics[width=15cm]{averageexposureonly.pdf} 
\caption{Average number of storm exposures per decade in U.S. counties for each 
	exposure metric. The criteria behind each of the five metrics is given in 
	Table \ref{tab:exposuremetrics}. The years used to estimate these averages 
	are based on years of available exposure data 
	(distance and wind:~1988\,--\,2015; rain: 1988\,--\,2011; 
	flood and tornado events:~1996\,--\,2015). } 
\label{fig:averageexposure} 
\end{figure}

@@

\begin{figure}%[tbhp]
\centering
\includegraphics[width=16cm]{ivanonly.pdf}
\caption{Counties classified as exposed to Hurricane Ivan in 2004 under each
exposure metric considered (Table~\ref{tab:exposuremetrics}). The red line
shows the track of Hurricane Ivan based on the revised Atlantic hurricane
database \ac{HURDAT2}~\parencite{landsea2013}.  Similar maps for other
large-extent storms are given in Figure~S3.}
\label{fig:ivanexposure} 
\end{figure}

@@

\begin{figure}%[tbhp] 
\centering 
\includegraphics[width = 0.6\linewidth]{jaccard_heatmap.pdf} 
\caption{Agreement between exposure classifications based on different
         exposure metrics for all storms between~1996 and~2011 for which 
	 at least~250 counties were exposed based on at least one metric.
	 Each row shows one storm, and the color of each cell shows the 
	 measured Jaccard index for each pair of exposure metrics 
	 (proportion of counties classified as exposed by both metrics out 
	 of storms classified as exposed by either metric). The colors to the 
	 right of the main heatmap for each storm indicate the total number of 
	 counties classified as exposed to the storm by any of the five metrics, 
	 providing an indication of each storm's extent. Storms are displayed 
	 within clusters that have similar patterns in county-level exposure 
	 agreement for metric pairs, based on hierarchical clustering using the 
	 complete link method~\parencite{murtagh2012algorithms}; columns are also 
	 ordered based on hierarchical clustering. Maps are available showing the 
	 counties identified as exposed under each of five metrics for the widest-extent 
	 storm in each cluster: Hurricane Ivan in~2004 (Figure~\ref{fig:ivanexposure}) 
	 and Hurricanes Floyd in~1999, Lee in~2011, Cindy in~2005, and Katrina 
	 in~2005 (Figure~S3).
} 
\label{fig:jaccard}
\end{figure}

>>>>

\begin{figure}[tbhp!] \centering
\includegraphics[width=0.7\linewidth]{hurrtracks}
\caption{States and storms considered in this study. All counties in the states
	shown in this map were investigated. The lines show the paths of the
	study storms, which included all tracked storms in 1988\,--\,2015 that
	are recorded in \ac{HURDAT2}~\parencite{landsea2013} and that came
	within 250~\si{\kilo\metre} of at least one \ac{US} county. Thicker
	lines show the tracks of storms whose names have been retired,
	indicating that the storm was particularly severe or had notable
	impacts~\parencite{retirednames}.  }
\label{fig:hurrtracks}
\end{figure}

@@

\begin{figure}[tbhp!] \centering
\includegraphics[width=0.9\linewidth]{raincomparison} 
	\caption{Comparison of storm-associated rainfall estimates between 
	a re-analysis dataset (included as the rainfall exposure metric
	in the open-source data and included in further analysis) and ground-based
	observations in nine sample counties. The rainfall estimates  
	include rainfall from two days before to one day after the storm's closest
	approach to the county. Each small graph shows data for one sample county, 
	and each point shows one tropical storm. The number of storms 
	within each county and the number of stations reporting rainfall during 
	the county's storms are given above each plot. Horizontal and vertical lines 
	in each plot show the threshold of 75~\si{\milli\metre} used to classify a storm 
	as ``exposed'' in further analysis (Table 1). Note that 
	ranges of the x and y axes differ across counties.
	} 
\label{fig:raincomparison}
\end{figure}

\begin{figure}[tbhp!]
\centering
\includegraphics[width=0.6\linewidth]{windcomparison}
\caption{Comparison of modeled maximum sustained surface wind speeds (included
	as the primary wind exposure metric in the open-source data and used in
	further analysis) with estimates from \ac{HURDAT2}'s wind speed radii.
	Each point represents a storm, with the x axis giving the percent of
	counties classified in the same category of maximum sustained surface
	wind speeds ($<$34~\si{\knot}; 34\,--\,49.9~\si{\knot};
	50\,--\,63.9~\si{\knot}; $\ge$64~\si{\knot}) by both sources of wind
	data for that storm. The color of each point indicates the extent of
	the storm, giving the number of study counties that were exposed to
	maximum sustained surface wind speeds of at least 34~\si{\knot} (based
	on modeled wind speeds). Estimates are shown for study storms since
	2004, the earliest year for which post-storm reanalysis wind speed
	radii estimates are routinely available in \ac{HURDAT2}, for which at
	least one study county had a sustained wind speed of $\ge$34~\si{\knot}
	based on the \ac{HURDAT2} wind radii.
	}
\label{fig:windcomparison}
\end{figure}

\begin{figure}[tbhp!]
\centering
\includegraphics[width=0.8\linewidth]{floodcomparison}
\caption{Comparison for a sample of study counties of flood status based on
	matching NOAA Storm Events listings to \ac{HURDAT2} storm track data
	versus five-day total streamflows at county streamflow gages during
	tropical cyclones. Each small plot shows results for one of the sample
	counties. Each point represents a single tropical storm, and the
	point's position along the x-axis shows the highest daily total
	streamflow (cubic feet per second), summed across all identified
	streamgages in the county, for a five-day window centered on the day of
	the storm's closest approach to the county. The y-axis separates storms
	for which a flood event was reported in NOAA's Storm Events database
	for the county with a start date within the five-day window of the
	storm's closest approach to the county. The color of each point gives
	the percent of streamflow gages in the county with a daily streamflow
	that exceeded a threshold of flooding (the streamgage's median value
	for annual peak flow) on any day during the five-day window. The number
	of streamflow gages used in analysis are given in parentheses beside
	the county's name in the panel title.  Note that the x-axis scales
	differ by county, depending on the number of streamflow gages and
	typical flow rates for each gage, and are on a log-10 scale.
	}
\label{fig:floodcomparison}
\end{figure}

@@

\begin{figure}%[tbhp] 
\centering
\includegraphics[width=7cm]{averageexposureonly.pdf} 
\caption{Average number of county-level storm exposures per decade for each
	single-hazard exposure metric. The criteria behind each of the five
	metrics is given in Table \ref{tab:exposuremetrics}. The years used to
	estimate these averages are based on years of available exposure data
	(rain: 1988\,--\,2011; wind:~1988\,--\,2015; flood and
	tornado events:~1996\,--\,2015). } 
\label{fig:averageexposure} 
\end{figure}

@@

\begin{figure}%[tbhp]
\centering
\includegraphics[width=16cm]{ivanonly.pdf}
\caption{Counties classified as exposed to Hurricane Ivan in 2004 under each
	exposure metric considered (Table~\ref{tab:exposuremetrics}). The red
	line shows the track of Hurricane Ivan based on
	\ac{HURDAT2}~\parencite{landsea2013}.  Similar maps for other
	large-extent storms are given in Figure~S4.
	}
\label{fig:ivanexposure} 
\end{figure}

@@

\begin{figure}%[tbhp] 
\centering 
\includegraphics[width = 0.6\linewidth]{jaccard_heatmap.pdf} 
\caption{Agreement between exposure classifications based on different
         exposure metrics for all storms between~1996 and~2011 for which 
	 at least~250 counties were exposed based on at least one metric.
	 Each row shows one storm, and the color of each cell shows the 
	 measured Jaccard index for each pair of exposure metrics 
	 (proportion of counties classified as exposed by both metrics out 
	 of storms classified as exposed by either metric). The colors to the 
	 right of the main heatmap for each storm indicate the total number of 
	 counties classified as exposed to the storm by any of the five metrics, 
	 providing an indication of each storm's extent. Storms are displayed 
	 within clusters that have similar patterns in county-level exposure 
	 agreement for metric pairs, based on hierarchical clustering using the 
	 complete link method~\parencite{murtagh2012algorithms}; columns are also 
	 ordered based on hierarchical clustering. Maps are available showing the 
	 counties identified as exposed under each of five metrics for the widest-extent 
	 storm in each cluster: Hurricane Ivan in~2004 (Figure~\ref{fig:ivanexposure}) 
	 and Hurricanes Floyd in~1999, Lee in~2011, Cindy in~2005, and Katrina 
	 in~2005 (Figure~S4).
} 
\label{fig:jaccard}
\end{figure}

### 2019-10-09 writing/ehp_introduction.tex
Work on paper, use 34 knot threshold

Tropical cyclones, including hurricanes, can have disastrous impacts in the
\acf{US}. A tropical cyclone's high winds can bring health risks
and property damage through structural damage of houses and other buildings,
falling trees, and wind-borne debris~\parencite{rappaport2000} and cause power
outages~\parencite{liu2005, han2009}, which introduce a number of threats to human
and ecological health, including water quality risks if the outage affects
wastewater treatment plants~\parencite{mallin2006}.  Other risks can exist without
severe wind; for example, one study found that most of the direct
hurricane-related deaths in the \ac{US} between~1970 and~1999 occurred in cases
when wind was below hurricane strength, including for Tropical Storms Charley
in~1998 and Alberto in~1994~\parencite{rappaport2000}.  Tropical cyclones can
produce excessive rain, especially in certain topographies (e.g., near
mountains), so counties well inland sometimes experience more extreme rain than
coastal counties. Flood risks from tropical cyclones can result from this rain,
although the two risks are not perfectly correlated~\parencite{chen2015}. In the
\ac{US}, over half of hurricane-related direct deaths from~1970 to~1999 from
Atlantic basin storms were a result of freshwater flooding~\parencite{rappaport2000}. 
Flooding can also degrade water quality~\parencite{mallin2006}, which can threaten both 
human and ecological health.

To measure tropical cyclone exposure for either investigations of exposure
patterns or for estimation of tropical cyclone risks and impacts, including
those related to human health, studies have varied widely in how they determine
tropical cyclone exposure (some examples shown in Figure~S1\todo{Check
fig.ref}). While some studies have measured exposure based on specific hazards
of the tropical cyclone (e.g., wind, rain), many have used distance from the
tropical cyclone's central track as a surrogate in identifying exposure to the
hazards of a tropical cyclone (e.g.,~\textcite{czajkowski2011, tansel2010,
kinney2008, caillouet2008increase}).

Distance from the tropical cyclone's central track can be easily measured using
widely-available hurricane tracking data. There are, however, a number of
limitations to using distance to assess exposure to tropical cyclones, and
there are reasons to suspect that exposure classifications based on other
tropical cyclone hazards may disagree with distance-based classifications.
Tropical cyclones vary dramatically in size: \ac{US} tropical cyclones have
been observed with radii to maximum winds as small as~20~\si{\kilo\metre} and as
large as~200~\si{\kilo\metre}~\parencite{mallin2006, quiring2011variations}. While a
number of tropical cyclone hazards are strongly associated with distance from
the tropical cyclone's center (e.g., wind and, at the coast, storm surge and
waves~\parencite{rappaport2000, kruk2010}), other hazards like dangerous rain,
floods, and tornadoes can occur well away from the tropical cyclone's central
track~\parencite{rappaport2000, atallah2007, moore2012}.  For example, fatal
tropical cyclone tornadoes, which were linked to over~300 deaths in the \ac{US}
between~1995 and~2009, most often occur~200\,--\,500~\si{\kilo\metre} from the
tropical cyclone's center~\parencite{moore2012}.

Further, distance-based exposure metrics that use
buffers tend to use an equal buffer distance on each side of the tropical
cyclone track (e.g.,~\cite{czajkowski2011, grabich2015, grabich2016,
zandbergen2009, tansel2010}), but the forces of a tropical cyclone tend to be
distributed around the center in a non-symmetrical way. For example, extreme
winds are more common to the right of the track, where counter-clockwise
cyclonic winds move in concert with the tropical cyclone's forward 
motion~\parencite{halverson2015}, and the fatal tornadoes associated with \ac{US} tropical
cyclones between~1995 and~2009 occurred almost exclusively to the right of the
tropical cyclone's track, mostly in the right front quadrant of the tropical
cyclone~\parencite{moore2012}. Rain, conversely, is often heaviest to the left of
the tropical cyclone's track, especially when the tropical cyclone interacts
with other weather systems~\parencite{atallah2003, atallah2007, zhu2013variations}
or undergoes an extratropical transition~\parencite{elsberry2002}.

\todo[inline]{Something about difference in other metrics}

The use of different exposure metrics across studies hampers the development of
a scientific consensus on tropical cyclone-related risks, as differences
observed between studies could result from differences in exposure assessment.
Further, if a study misclassifies exposure to the hazard or hazards that cause
an impact, estimates of tropical cyclone risks or impacts will be biased,
making it hard to identify true associations. These concerns are particularly
important if there are large differences in the locations that are determined
to be exposed based on different metrics.

To investigate this, we measured county-level exposure to tropical cyclones in
all eastern \ac{US} counties using five metrics
(Table~\ref{tab:exposuremetrics}): (1)~distance to tropical cyclone track;
(2)~maximum sustained wind speed; (3)~cumulative rainfall; (4)~flood events;
and (5)~tornado events.  We assessed exposure at the county level since data on
many potential impacts are available at county-level aggregations (e.g., direct
hurricane-related deaths~\parencite{czajkowski2011}; birth
outcomes~\parencite{grabich2015, grabich2016}; autism
prevalence~\parencite{kinney2008}) and since decisions and policies to prepare for,
and respond to, tropical cyclones are often undertaken at the county
level~\parencite{zandbergen2009, rappaport2000}.  We explored patterns in these
exposure assessments and measured how well exposure classification agreed
across these five metrics in terms of classifying specific counties as exposed
to a tropical cyclone.  Finally, to make this hazard-specific tropical cyclone
exposure data available to other researchers, we published open source software
for county-based hurricane exposure assessment in \ac{US}
counties~\parencite{hurricaneexposure}.

>>>>

\acresetall

Recently, several major tropical cyclones have hit \ac{US} cities, including
Hurricanes Harvey, Irma, and Maria in 2017 \parencite{blake20182017} and
Hurricanes Florence and Michael in 2018 \parencite{avila20192018}. Researchers
are exploring how these storms impacted human health (e.g.,
\parencite{santos2018use, rivera2018estimating, santos2018differential,
grineski2019impact, issa2018deaths, tanz2019notes, paul2019brief}), including
through projects funded through \ac{NIH} Rapid Response grants
\parencite{nihreporter}. These projects will add to evidence from previous
hurricanes, including Sandy in 2012 (e.g., \parencite{swerdel2014}) and Katrina
in 2005 (e.g., \parencite{burton2009health}), characterizing the health impacts
of hurricanes.

A recent review highlights the need to supplement these single-storm studies
with multi-year, multi-storm studies, to identify patterns that are common
across multiple storms [ref]. While multi-year studies of hurriance mortality
impacts have been conducted using disaster surveillance data, studies based on
administrative data could identify risks for non-accidental mortality and
morbidity that those studies likely undercount [ref]. Multi-storm studies with
administrative data require exposure assessment that is consistent across time
and space, allowing the administrative data to be linked to disaster exposure
[ref]. Key to expanding tropical cyclone epidemiology are therefore exposure
assessment methods that appropriately capture hazards of the storm that are
relevant to human health, methods that can be applied consistently across
multiple storms, locations, and years.

The \ac{NHC} publishes a ``Best Tracks'' dataset that is considered the gold
standard for Atlantic-basin tropical cyclone tracks.  It records the central
position of a tropical cyclone every six hours, as well as the storm's
\underline{minimum pressure} and \underline{central maximum sustained winds}. This data is openly
available through the \ac{HURDAT2}, a post-storm assessment conducted
by the \ac{US} \ac{NHC} that incorporates data from a variety of sources,
including satellite data and, when available, aircraft reconnaissance
data~\parencite{landsea2013, jarvinen1988}. With this data, it is
straightforward to measure whether a community was in the direct path, or
within a certain distance of the path, of a storm. Some studies have indeed
used this data to assess tropical cyclone exposure based on how near the
storm's central track came to the county, either to assess average exposure
patterns to tropical storms [refs] or to characterize exposure for health
studies [refs].

When epidemiological studies assess exposure to a tropical cyclone based on 
how close the storm came to a community, however, they may misclassify exposure. 
Tropical cyclones vary dramatically in size: \ac{US} tropical
cyclones have been observed with radii to maximum winds as small
as~20~\si{\kilo\metre} and as large
as~200~\si{\kilo\metre}~\parencite{mallin2006, quiring2011variations}.  While a
number of tropical cyclone hazards are strongly associated with distance from
the tropical cyclone's center (e.g., wind and, at the coast, storm surge and
waves~\parencite{rappaport2000, kruk2010}), other hazards like heavy rainfall,
floods, and tornadoes can occur well away from the tropical cyclone's central
track~\parencite{rappaport2000, atallah2007, moore2012}.  For example, fatal
tropical cyclone tornadoes, which were linked to over~300 deaths in the \ac{US}
between~1995 and~2009, most often occur~200\,--\,500~\si{\kilo\metre} from the
tropical cyclone's center~\parencite{moore2012}.

Further, when studies use distance from the storm's track to assess exposure,
they often use an equal buffer distance on each side of the tropical cyclone
track (e.g.,~\cite{czajkowski2011, grabich2015, grabich2016, zandbergen2009,
tansel2010}). However, the forces of a tropical cyclone tend to be distributed
around the center in a non-symmetrical way. Extreme winds are more common to
the right of the track, where counter-clockwise cyclonic winds move in concert
with the tropical cyclone's forward motion~\parencite{halverson2015}, and the
fatal tornadoes associated with \ac{US} tropical cyclones between~1995 and~2009
occurred almost exclusively to the right of the tropical cyclone's track,
mostly in the right front quadrant of the tropical
cyclone~\parencite{moore2012}. Rain, conversely, is often heaviest to the left
of the tropical cyclone's track, especially when the tropical cyclone interacts
with other weather systems~\parencite{atallah2003, atallah2007,
zhu2013variations} or undergoes an extratropical
transition~\parencite{elsberry2002}.

If a study misclassifies exposure to the hazard or hazards that cause the
health risk being studied, the study will generate biased estimates of tropical
cyclone risks and impacts. Further, when different studies use different
methods to assess exposure to tropical cyclones, their results are difficult to
meaningfully compare and aggregate. If different methods identify
similar sets of communities as ``exposed'',  these concerns are less serious.
However, if different methods differ substantially in which communities they
identify as ``exposed'', it makes it very important that epidemiological
studies are thoughtful in how they assess exposure.

To provide open-source data on tropical cyclone exposures to other researchers,
we developed a dataset of county-level exposure to tropical cyclones in all
eastern \ac{US} counties for five different metrics
(Table~\ref{tab:exposuremetrics}): (1)~distance to tropical cyclone track;
(2)~maximum sustained wind speed; (3)~cumulative rainfall; (4)~flood events;
and (5)~tornado events~\parencite{hurricaneexposure}. This data is provided at
the county level since data on many potential impacts are available at
county-level aggregations (e.g., direct hurricane-related
deaths~\parencite{czajkowski2011}; birth outcomes~\parencite{grabich2015,
grabich2016}; autism prevalence~\parencite{kinney2008}) and since decisions and
policies to prepare for, and respond to, tropical cyclones are often undertaken
at the county level~\parencite{zandbergen2009, rappaport2000}. Further, we
explored patterns in these exposure assessments across storm hazards and
measured how well exposure classification agreed across these five metrics in
terms of classifying specific counties as exposed to a tropical cyclone.
